---
format:
  pdf:
    include-in-header: 
      - text: |
            \usepackage{amsmath}
execute: 
  echo: false
  warning: false
  message: false
editor: visual
---

```{=latex}
\begin{titlepage}
\begin{center}
\vspace{2 cm}

{\Large STATISTICS HONOURS - LIKELIHOOD}\\
\vspace{0.3cm}
{\large Department of Statistical Sciences\\University of Cape Town}\\
\vspace{6 cm}

\rule{16 cm}{0.3 mm}\\
\vspace{0.5 cm}
{\Large \textbf{Assignment 2}}\\
\vspace{0.5cm}
\rule{16 cm}{0.3 mm}\\

\vspace{2 cm}

\begin{minipage}{0.4\textwidth}
\begin{flushleft}\large
\emph{Authors:}\\
\vspace{0.5 cm}
Natalie Chidikwindi\\
Rosie Chen\\
Donnica Haupt
\end{flushleft}
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{flushright}\large
\emph{Student Numbers:}\\
\vspace{0.5 cm}
CHDNAT002\\
CHNYOJ001\\
HPTDON002
\end{flushright}
\end{minipage}

\vfill
{\large \today}\\
\end{center}
\end{titlepage}

\newpage
```

```{r setup}
# loading libraries
library(tidyverse)
library(cmdstanr)
library(knitr)
```

```{r dataset}
coffeetime <- read.csv("coffeetime.csv")
N <- nrow(coffeetime) # sample size
n <- n_distinct(coffeetime$cafe) # number of cafes
t <- coffeetime$time # waiting time
i <- coffeetime$cafe # cafe index
dat <- list(N = N, n = n, t = t, i = i)
```

# Question 1:

To express the integrand

$$L(\boldsymbol\theta) = \pi(\textbf{t} \mid \boldsymbol\theta) = \int \pi (\textbf{t} \mid \boldsymbol\lambda) \pi (\boldsymbol\lambda \mid \boldsymbol\theta) d \boldsymbol\lambda$$

in Equation 2 as a product of $n$ functions, each of which depends on only one $\lambda_i$, we first need to factorise the conditional likelihood $\pi (\textbf{t} \mid \lambda)$ (the first part of the integrand).

Since $T_i \sim$ Exponential$(\lambda_i)$, this means that the waiting time $t_i$ at café $i$ depends only on its specific rate parameter $\lambda_i$. Therefore, taking the conditional on $\lambda$, the individual observations $t_i$ are conditionally independent.

Expressing this joint distribution as the product of individual distributions we have $$\prod_{i=1}^n \pi(t_i \mid \lambda_i).$$

The second part of the integrand is $\pi(\boldsymbol\lambda \mid \boldsymbol\theta)$. This is the joint prior distribution of all $n$ rate parameters, given the shared hyperparameters $\boldsymbol\theta = (\alpha, \beta)$.

Since it is stated in Equation 1 that $\lambda_i \overset{\mathrm{iid}}{\sim}$Gamma($\alpha, \beta$) for $i= 1, ... n$. This implies that the parameters $\lambda_1,...,\lambda_n$ are independent and identically distributed (i.i.d.) conditional on the hyperparameters $\theta$. Hence, we can express the joint prior as a product of individual priors as shown below

$$\prod_{i=1}^n \pi(\lambda_i \mid \theta).$$ Substituting the two products that we found earlier back into the integrand we have,

$$\pi(\textbf{t} \mid \boldsymbol\lambda) \pi(\boldsymbol\lambda \mid \boldsymbol\theta) = \left( \prod_{i=1}^n \pi(t_i \mid \lambda_i) \right) \left( \prod_{i=1}^n \pi(\lambda_i \mid \theta) \right).$$

Since the products iterate over the same index $i$, we can group the terms for each $i$ together, which gives us

$$\prod_{i=1}^{n} \Big[ \pi(t_{i}\mid \lambda_{i})\pi(\lambda_{i} \mid \theta) \Big].$$

Now, substituting the rearranged integrand back into the full integral we have the following:

$$L(\theta) = \int \dots \int \prod_{i=1}^{n} \Big[ \pi(t_{i}|\lambda_{i})\pi(\lambda_{i}|\theta) \Big] d\lambda_i \dots d\lambda_{n}.$$

Since each factor $\Big[ \pi(t_{i}|\lambda_{i})\pi(\lambda_{i}|\theta) \Big]$ depends on $\lambda$ only through $\lambda_i$, by Fubini's theorem the $n$-dimensional integral separates into a product of $n$ one-dimensional integrals.

Hence, we can pull the product operator outside of the integral and express the likelihood of $\theta$ as

$$L(\theta) = \prod_{i=1}^{n} \int \pi(t_{i}|\lambda_{i})\pi(\lambda_{i}|\theta) d\lambda_i, $$ which is a product of $n$ functions, each of which depends on only one $\lambda_i$ as required.

# Question 3

Since repeated visits can take place , let $t_{ij}$ represent the waiting time for coffee j at cafe i. Therefore the model specifications will be:

$$T_{ij} \mid \lambda_i {\sim}\text{Exponential}(\lambda_i)$$ where: $$ i = 1,2,3,\dots,n \quad \& \quad j = 1,2,3,\dots,v_i$$ $$\lambda_i \overset{\mathrm{iid}}{\sim} \text{Gamma}(\alpha, \beta)$$

# Question 4

Following the same logic as Question 1, but now accounting for $v_i$ visits per café, we marginalise out $\lambda_i$ from the joint distribution. Conditional on $\lambda_i$, the $v_i$ waiting times at caf'e $i$ are independent, so the joint density within café $i$ factorises as $\prod_{j=1}^{v_i} \pi(t_{ij} | \lambda_i)$. Since the caf'es are also conditionally independent given $\boldsymbol{\lambda}$, the full likelihood becomes:

$$L_i(\alpha, \beta | t_i) = \int\prod_{j=1}^{v_i} \pi(t_{ij}  | \lambda_i)  \pi(\lambda_i | \alpha, \beta) d\lambda_i$$

where $t_{ij}$ is the waiting time for visit $j$ at café $i$.

Our rate parameter $\lambda_i$ remains unchanged---the waiting time distribution for each café is independent of how many visits we make. Since $\lambda_i$ is an unobserved random variable, we integrate it out using its distribution $\pi(\lambda_i | \alpha, \beta)$, where $\alpha$ and $\beta$ are the fixed parameters we're estimating.

Since the waiting times across cafés are independent, we obtain the full likelihood by taking the product of the marginal likelihoods for each café:

$$L(\alpha, \beta \mid t_1, t_2, \dots, t_n) = \prod_{i = 1}^n \left[ \int \prod_{j=1}^{v_i} \pi(t_{ij} \mid \lambda_i) \pi(\lambda_i \mid \alpha, \beta) \, d\lambda_i \right]$$

# Question 5:

Given $\lambda_i \overset{iid}{\sim} \text{Gamma}(\alpha,\beta)$ for $i=1,\dots,n$ from question 3. $\lambda_i$ is the mean waiting time for a coffee at café $i$ with a Gamma($\alpha$,$\beta$) distribution. The distribution has a mean and variance of $\mu=\frac{\alpha}{\beta}$ and $\sigma^2=\frac{\alpha}{\beta^2}$, respectively. Based on our domain knowledge, a reasonable prior guess is that the mean waiting time for a coffee at a café is 5 minutes and the standard deviation between café's waiting times are 3 minutes. By solving for $\beta$ we have $$\beta=\frac{\mu}{\sigma^2}=\frac{5}{9}\approx0.56$$ thus $$\alpha = \mu\beta=\frac{25}{9}\approx2.78.$$

Exponential distributions have maximum entropy among distributions over positive real numbers with finite expectations. Therefore, let $\alpha$ and $\beta$ be exponential priors with rate parameters. Resulting in the following full Bayesian specification for the model in question 3: $$\begin{split} T_{ij} &\sim \text{Exponential}(\lambda_i),\\ \lambda_i &\overset{iid}{\sim} \text{Gamma}(\alpha,\beta), \\ \alpha &\sim \text{Exponential}(0.36), \\ \beta &\sim \text{Exponential}(1.8) \end{split}$$ where $i=1,\dots,n$ and $j=1,\dots,v_i$.

# Question 6:

The model in Question 5 was implemented in Stan and fitted to the $\texttt{coffeetime}$ data using four chains with 1000 warmup and 2000 sampling iterations with seed of 123.

```{r output=FALSE}
re_stan_code <- "
data {
  int<lower=0> N;
  int<lower=0> n;
  vector<lower=0>[N] t;
  array[N] int<lower=1, upper=n> i;
}

parameters {
  real<lower=0> alpha;
  real<lower=0> beta;
  vector<lower=0>[n] lambda;
}

model {
  alpha ~ exponential(0.36);
  beta ~ exponential(1.8);
  lambda ~ gamma(alpha, beta);
  for (k in 1:N) t[k] ~ exponential(1.0/lambda[i[k]]);
}
"

re_mod <- cmdstan_model(write_stan_file(re_stan_code)) # compile C++ model from Stan code

# sample posterior with MCMC
re_fit <- re_mod$sample(
  data            = dat, 
  seed            = 123,
  refresh         = 0,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 2000
)
```

```{r}
#| label: tbl-modsum
#| tbl-cap: "Model Summary"
pars_re = c("alpha", "beta", "lambda")
mod_sum <- re_fit$summary(variables = pars_re)
kable(mod_sum, digits = 3)
```

The model summary is given in @tbl-modsum which shows that all $\hat{R}$ values are approximately 1 and all ESS values are large. This indicates there was good convergene and efficient mixing. The posterior estimates for $\alpha$ and $\beta$ have a mean of `r round(mod_sum[1,2],2)` and `r round(mod_sum[2,2],2)` respectively, with the mean of $\lambda_i$ ranging from `r round(min(mod_sum[3:17,2]),2)` to `r round(max(mod_sum[3:17,2]),2)` minutes.

```{r}
#| label: tbl-diagsum
#| tbl-cap: "Model Diagnostic Summary"
cbind(1:4,as.data.frame(re_fit$diagnostic_summary())) |>
  kable(col.names = c("chains","divergences","max treedepth","EBFMI"), digits = 3)
```

The MCMC diagnostics confirmed that there was reliable sampling. The diagnostics summary is given in @tbl-diagsum and shows that there was no divergences or maximum tree depths that were hit and the E-BFMI is above 0.3 across all chains.

```{r}
#| label: fig-modhist
#| fig-cap: "Posterior distribution of mean waiting time across cafés"
re_post <- re_fit$draws(variables = pars_re, format = "df") |>
  as_tibble() |>
  select(-starts_with(".")) |>
  mutate(mu = alpha/beta)

ggplot(re_post, aes(x = mu)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  labs(
    x = expression("Mean expected waiting time "*(mu)),
    y = "Count"
  ) +
  theme_minimal()
```

The histogram in @fig-modhist displays the mean expected waiting time across cafés, where $\mu=\frac{\alpha}{\beta}$. The histogram is right skewed with the data focused between 3 and 7 minutes, with a mean of `r round(mean(re_post$mu),2)` which is similar to the observed sample mean of `r round(mean(coffeetime$time),2)` minutes.

# Question 7:

In Question 5, the "random effects" model stated that the café-specific means shared a common distribution of $$ \lambda_i \overset{\mathrm{iid}}{\sim} \text{Gamma}(\alpha, \beta).$$

This means that information is pooled across the cafés, which indicates that the estimate for café $i$ is influenced by what is observed across all cafés (through $\alpha \text{ and } \beta$).

To specify a Bayesian "fixed effect" model, we treat each $\lambda_i$ as a separate, independent parameter with no shared distribution, this means that each $\lambda_i$ now gets its own prior with no connection (information) to other cafés.

Since $\lambda_i$ is the mean waiting time, it must be positive. As in Question 5, an Exponential prior is used.

So we have $$T_{ij} \mid \lambda_i \sim \text{Exponential}(\lambda_i)$$ $$\lambda_i \sim \text{Exponential}(\phi), \quad i = 1,...,n$$ where $\phi$ is a fixed hyperparameter.

To find the value of $\phi$, we used our domain knowledge about café waiting times.

Since $\lambda_i \sim \text{Exponential}(\phi)$, the mean of the prior is: $$ E[\lambda_i] = \frac{1}{\phi}.$$

A reasonable guess is that the average waiting time at a café is 5 minutes, this means that setting $\frac{1}{\phi} = 5$ gives $\phi = 0.2$.

Hence, we have $$T_{ij} \mid \lambda_i \sim \text{Exponential}(\lambda_i)$$ $$\lambda_i \sim \text{Exponential}(0.2)$$ where $i = 1,...,n$ and $j=1,\dots,v_i$.

# Question 8:

```{r output=FALSE}
fe_stan_code <- "
data {
  int<lower=0> N;
  int<lower=0> n;
  vector<lower=0>[N] t;
  array[N] int<lower=1, upper=n> i; 
}
parameters {
  vector<lower=0>[n] lambda;                
}
model {
  lambda ~ exponential(0.2); 
  for (k in 1:N) t[k] ~ exponential(1.0/lambda[i[k]]);
}
"

fe_mod <- cmdstan_model(write_stan_file(fe_stan_code))

fe_fit <- fe_mod$sample(
  data            = dat,
  seed            = 123,
  refresh         = 0,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 2000
)
```

```{r}
#| label: tbl-modsum2
#| tbl-cap: "Model Summary"

pars_fe = "lambda"
mod_sum2 <- fe_fit$summary(variables = pars_fe)
kable(mod_sum2, digits = 3)
```

The $\hat{R}$ values for all lambdas shown in @tbl-modsum2 are extremely close to one, ranging from 0.9998 to 1.0019, this means that all 4 chains successfully converged to the same distribution. The ess_bulk values are all \> 15000, and ess_tail values range from approximately 5000 to 6000. These values are very high, indicating efficient sampling. The posterior estimate for $\lambda_i$ have a mean that ranges from `r round(min(mod_sum2[1:15,2]),2)` to `r round(max(mod_sum2[1:15,2]),2)`minutes.

```{r}
#| label: tbl-diagsum2
#| tbl-cap: "Model Diagnostic Summary"

cbind(1:4,as.data.frame(fe_fit$diagnostic_summary())) |>
  kable(col.names = c("chains","divergences","max treedepth","EBMFI"), digits = 3)
```

@tbl-diagsum2 shows that the divergence is 0 across all chains. This means that the model is well-behaved and did not wander off the posterior surface. No transitions hit the maximum treedepth across any chain, indicating that the sampler had sufficient room to fully explore the posterior at each step without being forced to terminate early. This means that the posterior is not very difficult to traverse and the sampler did not need many steps to explore the regions thoroughly. All EBFMI values are very close to 1. This implies that the sampler's energy transitions are explaining nearly all the posterior variability, which means all regions of the posterior have been explored by the sampler.

# Question 9

```{r}
#| label: fig-comparison
#| fig-cap: "Posterior medians and 90% credible intervals for café mean waiting times"

compare_df <- rbind(mod_sum[-c(1,2), ],mod_sum2) |>
  mutate(model = c(rep("Random effects", 15), rep("Fixed effects", 15)), 
         cafe = c(1:15,1:15))

ggplot(compare_df, aes(x = cafe, y = median, colour = model, 
                       ymin = q5, ymax = q95)) +
  geom_hline(yintercept = median(re_post$mu), lty = "dashed", colour = "grey40") +
  geom_errorbar(position = position_dodge(width = 0.5), width = 0.3) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_x_continuous(breaks = 1:15) +
  scale_colour_manual(values = c("Random effects" = "steelblue",
                                 "Fixed effects" = "salmon")) +
  labs(x = "Café", y = expression("Expected waiting time "*mu*" (minutes)"),
       colour = "Model") +
  theme_minimal()
```

As shown in the plot, the random effects estimates are being pulled towards the posterior median of the mean expected waiting time across cafés, represented by the dashed horizontal line, in comparison to the fixed effects estimates which are further from it, which is most visible at the extremes, such as cafés 8 and 10. This makes sense intuitively, as the random effects model uses shrinkage to share information across all cafés rather than estimating each one completely independently. For most of the cafés, the credibility intervals from the random effects model are narrower and more stable compared to those from the fixed effects model.

# Question 10

The population parameters $\alpha$ and $\beta$ define the Gamma distribution from which each café's mean waiting time $\lambda_i$ is drawn. Information about $\alpha$ and $\beta$ therefore accumulates through observing many cafés. Each café provides one independent draw from this population distribution.

As the number of visits to Café Statistique grows ($v \to \infty$), the posterior for that café's $\lambda_i$ becomes increasingly precise, eventually collapsing to a point. However, this does not increase $n$. There are still only 15 cafés providing information about the population parameters $\alpha$ and $\beta$, with 14 of them remaining estimated with uncertainty.

From Fisher information theory, information about $(\alpha, \beta)$ accumulates with the number of independent cafés $n$, not with the number of visits to a single café. Since $n = 15$ is fixed, the total Fisher information is bounded. By the Cramér-Rao lower bound, the standard error of $\mu = \alpha / \beta$ therefore has a positive lower bound it cannot cross, regardless of how many visits are made to Café Statistique. The standard error will decrease initially as $\lambda_{\text{Statistique}}$ becomes more precise, but will plateau at a positive non-zero value.
