---
title: "Question 3&4"
author: "Natalie C"
format: pdf
editor: visual
---

```{r}
#loading of libraries
library(readr)
library(tidyverse)
library(cmdstanr)
library(ggplot2)
library(knitr)
```

## Loading for dataset

```{r}
coffeetime <- read_csv("coffeetime.csv")
N <- nrow(coffeetime) # sample size
n <- n_distinct(coffeetime$cafe) # number of cafes
t <- coffeetime$time # waiting time
i <- coffeetime$cafe # cafe index
```

### Question 1:

To express the integrand

$$
L(\boldsymbol\theta) = \pi(\textbf{t} \mid \boldsymbol\theta) = \int \pi (\textbf{t} \mid \boldsymbol\lambda) \pi (\boldsymbol\lambda \mid \boldsymbol\theta) d \boldsymbol\lambda
$$

in Equation 2 as a product of $n$ functions, each of which depends on only one $\lambda_i$, we first need to factorise the conditional likelihood $\pi (\textbf{t} \mid \lambda)$ (the first part of the integrand).

Since $T_i \sim$ Exponential$(\lambda_i)$, this means that the waiting time $t_i$ at café $i$ depends only on its specific rate parameter $\lambda_i$. Therefore, taking the conditional on $\lambda$, the individual observations $t_i$ are conditionally independent.

Expressing this joint distribution as the product of individual distributions we have $$
\prod_{i=1}^n \pi(t_i \mid \lambda_i).
$$

The second part of the integrand is $\pi(\boldsymbol\lambda \mid \boldsymbol\theta)$. This is the joint prior distribution of all $n$ rate parameters, given the shared hyperparameters $\boldsymbol\theta = (\alpha, \beta)$.

Since it is stated in Equation 1 that $\lambda_i \overset{\mathrm{iid}}{\sim}$Gamma($\alpha, \beta$) for $i= 1, ... n$. This implies that the parameters $\lambda_1,...,\lambda_n$ are independent and identically distributed (i.i.d.) conditional on the hyperparameters $\theta$. Hence, we can express the joint prior as a product of individual priors as shown below

$$
\prod_{i=1}^n \pi(\lambda_i \mid \theta).
$$ Substituting the two products that we found earlier back into the integrand we have,

$$
\pi(\textbf{t} \mid \boldsymbol\lambda) \pi(\boldsymbol\lambda \mid \boldsymbol\theta) = \left( \prod_{i=1}^n \pi(t_i \mid \lambda_i) \right) \left( \prod_{i=1}^n \pi(\lambda_i \mid \theta) \right).
$$

Since the products iterate over the same index $i$, we can group the terms for each $i$ together, which gives us

$$
\prod_{i=1}^{n} \Big[ \pi(t_{i}\mid \lambda_{i})\pi(\lambda_{i} \mid \theta) \Big].
$$

Now, substituting the rearranged integrand back into the full integral we have the following:

$$
L(\theta) = \int \dots \int \prod_{i=1}^{n} \Big[ \pi(t_{i}|\lambda_{i})\pi(\lambda_{i}|\theta) \Big] d\lambda_i \dots d\lambda_{n}.
$$

Since the limits of integration are independent and the integrand is separable as a product of functions of distinct variables, the joint integral simplifies to the product of \$n\$ individual, one-dimensional integrals.

Hence, we can pull the product operator outside of the integral and express the likelihood of $\theta$ as

$$L(\theta) = \prod_{i=1}^{n} \int \pi(t_{i}|\lambda_{i})\pi(\lambda_{i}|\theta) d\lambda_i, 
$$ which is a product of $n$ functions, each of which depends on only one $\lambda_i$ as required.

### Question 3

Since repeated visits can take place , let $t_{ij}$ represent the waiting time for coffee j at cafe i. There fore the model specifications will be:

$$T_{ij} \overset{\mathrm{iid}}{\sim}\text{Exponential}(\lambda_i)$$ where: $$ i = 1,2,3,....n \quad \text{&} \quad j = 1,2,3,....,v_i$$ $$\lambda_i \overset{\mathrm{iid}}{\sim} \text{Gamma}(\alpha, \beta)$$ \* used LMM to be able to typeset it correctly (the iid bit at the \~ bit).

### Question 4

In Question 1 we have:

If we are told that the visits to the coffee shops are independent, then we can rewrite the likelihood to visit cafe i in the following way: $$ L_i(\alpha, \beta | t_i) = \int\prod_j^{v_i} \pi(t_{ij}  | \lambda_i)  \pi(\lambda_i | \alpha, \beta) d\lambda_i$$

Where $t_{ij}$ is the waiting time for coffee j at time i.

Our rate parameter remains unchanged as the waiting time for each cafe remains the same regardless how many times you go for another cup of coffee. Since $\lambda_i$ is an unobserved random variable, we integrate it out using its distribution $\pi(\lambda_i | \alpha, \beta)$. The parameters $\alpha$ and $\beta$ are the fixed parameters we're estimating.

Since the waiting times for the cafes are also independent we can take the product of the marginal likelihood for each cafe to find the general likelihood.

$$L(\alpha, \beta \mid t_1, t_2, \dots, t_n) = \prod_{i = 1}^n \left[ \int \prod_{j=1}^{v_i} \pi(t_{ij} \mid \lambda_i) \pi(\lambda_i \mid \alpha, \beta) \, d\lambda_i \right]$$

### Question 5:

Given $\lambda_i \overset{iid}{\sim} \text{Gamma}(\alpha,\beta)$ for $i=1,\dots,n$ from question 3. $\lambda_i$ is the mean waiting time for a coffee at café $i$ with a Gamma($\alpha$,$\beta$) distribution. The distribution has a mean and variance of $\mu=\frac{\alpha}{\beta}$ and $\sigma^2=\frac{\alpha}{\beta^2}$, respectively. Based on our domain knowledge, a reasonable prior guess is that the mean waiting time for a coffee at a café is 5 minutes and the standard deviation between café's waiting times are 3 minutes. By solving for $\beta$ we have $$\beta=\frac{\mu}{\sigma^2}=\frac{5}{9}\approx0.56$$ thus $$\alpha = \mu\beta=\frac{25}{9}\approx2.78.$$

Exponential distributions have maximum entropy among distributions over positive real numbers with finite expectations. Therefore, let $\alpha$ and $\beta$ be exponential priors with rate parameters. Resulting in the following full Bayesian specification for the model in question 3: $$\begin{split} T_{ij} &\sim \text{Exponential}(\lambda_i),\\ \lambda_i &\overset{iid}{\sim} \text{Gamma}(\alpha,\beta), \\ \alpha &\sim \text{Exponential}(0.36), \\ \beta &\sim \text{Exponential}(1.8) \end{split}$$ where $i=1,\dots,n$ and $j=1,\dots,v_i$.

### Question 6:

```{r, message=FALSE, output=FALSE}
stan_code_q6 <- "
data {
  int<lower=0> N;
  int<lower=0> n;
  vector<lower=0>[N] t;
  array[N] int<lower=1, upper=n> i;
}

parameters {
  real<lower=0> alpha;
  real<lower=0> beta;
  vector<lower=0>[n] lambda;
}

model {
  alpha ~ exponential(0.36);
  beta ~ exponential(1.8);
  lambda ~ gamma(alpha, beta);
  for (k in 1:N) t[k] ~ exponential(1.0/lambda[i[k]]);
}
"

stan_file_q6 <- write_stan_file(stan_code_q6)
mod_q6 <- cmdstan_model(stan_file_q6) # compile C++ model from Stan code
stan_data_q6 <- list(N = N, n = n, t = t, i = i)
# sample posterior with MCMC
fit_q6 <- mod_q6$sample(
  data = stan_data_q6, 
  seed = 123,
  refresh = 0,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 2000
)

pars = c("alpha", "beta", "lambda")
kable(fit_q6$summary(variables = pars))
cbind(1:4,as.data.frame(fit_q6$diagnostic_summary())) |>
  kable(col.names = c("chains","divergences","max treedepth","EBMFI"))

post_q6 <- fit_q6$draws(variables = pars, format = "df") |>
  as_tibble() |>
  select(-starts_with(".")) |>
  mutate(mu = alpha/beta)

ggplot(post_q6, aes(x = mu)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  labs(
    x = expression("Mean expected waiting time "*(mu)),
    y = "Count",
    title = "Posterior distribution of mean waiting time across cafés"
  ) +
  theme_minimal()
```

### Question 7:

In Question 5, the "random effects" model stated that the café-specific means shared a common distribution of $$ \lambda_i \overset{\mathrm{iid}}{\sim} \text{Gamma}(\alpha, \beta).$$

This means that information is pooled across the cafés, which indicates that the estimate for café $i$ is influenced by what is observed across all cafés (through $\alpha \text{ and } \beta$).

To specify a Bayesian "fixed effect" model, we treat each $\lambda_i$ as a separate, independent parameter with no shared distribution, this means that each $\lambda_i$ now gets its own prior with no connection (information) to other cafés.

Since $\lambda_i$ is the mean waiting time, it must be positive. Using information given in Question 5, we are going to use the Exponential distribution once again for the prior as the parameter is positive.

So we have $$T_{ij} \mid \lambda_i \sim \text{Exponential}(\lambda_i)$$ $$\lambda_i \sim \text{Exponential}(\phi), \quad i = 1,...,n$$ where $\phi$ is a fixed hyperparameter.

To find the value of $\phi$, we used our domain knowledge about café waiting times.

Since $\lambda_i \sim \text{Exponential}(\phi)$, the mean of the prior is: $$ E[\lambda_i] = \frac{1}{\phi}.$$

A reasonable guess is that the average waiting time at a café is 5 minutes, this means that setting $\frac{1}{\phi} = 5$ gives $\phi = 0.2$.

Hence, we have $$T_{ij} \mid \lambda_i \sim \text{Exponential}(\lambda_i)$$ $$\lambda_i \sim \text{Exponential}(0.2), \quad i = 1,...,n$$


### Question 8:
```{r}
stan_code_q8 <- "
data {
  int<lower=0> N;
  int<lower=0> n;
  vector<lower=0>[N] t;
  array[N] int<lower=1, upper=n> i; 
}
parameters {
  vector<lower=0>[n] lambda;                
}
model {
  lambda ~ exponential(0.2); 
  for (k in 1:N) t[k] ~ exponential(1.0/lambda[i[k]]);
}
"
mod <- cmdstan_model(write_stan_file(stan_code_q8))

stan_data <- list(N = N, n = n, t = t, i = i)

fit_q8 <- mod$sample(
  data            = stan_data,
  seed            = 123,
  chains          = 4,
  parallel_chains = 4,
  iter_warmup     = 1000,
  iter_sampling   = 2000
)

pars_q8 = "lambda"
kable(fit_q8$summary(variables = pars_q8))
cbind(1:4,as.data.frame(fit_q8$diagnostic_summary())) |>
  kable(col.names = c("chains","divergences","max treedepth","EBMFI"))
```
