---
title: "Question 3&4"
author: "Natalie C"
format: pdf
editor: visual
---

```{r}
#loading of libraries
library(readr)
```

## Loading for dataset

```{r}
coffeetime <- read_csv("coffeetime.csv")
```

### Question 3

Since repeated visits can take place , let $t_{ij}$ represent the waiting time for coffee j at cafe i. There fore the model specifications will be:

$$T_{ij} \overset{\mathrm{iid}}{\sim}\text{Exponential}(\lambda_i)$$ where: $$ i = 1,2,3,....n \quad \text{&} \quad j = 1,2,3,....,v_i$$ $$\lambda_i \overset{\mathrm{iid}}{\sim} \text{Gamma}(\alpha, \beta)$$ \* used LMM to be able to typeset it correctly (the iid bit at the \~ bit).

### Question 4

In Question 1 we have:

If we are told that the visits to the coffee shops are independent, then we can rewrite the likelihood to visit cafe i in the following way: $$ L_i(\alpha, \beta | t_i) = \int\prod_j^{v_i} \pi(t_{ij}  | \lambda_i)  \pi(\lambda_i | \alpha, \beta) d\lambda_i$$

Where $t_{ij}$ is the waiting time for coffee j at time i.

Our rate parameter remains unchanged as the waiting time for each cafe remains the same regardless how many times you go for another cup of coffee. Since $\lambda_i$ is an unobserved random variable, we integrate it out using its distribution $\pi(\lambda_i | \alpha, \beta)$. The parameters $\alpha$ and $\beta$ are the fixed parameters we're estimating.

Since the waiting times for the cafes are also independent we can take the product of the marginal likelihood for each cafe to find the general likelihood.

$$L(\alpha, \beta \mid t_1, t_2, \dots, t_n) = \prod_{i = 1}^n \left[ \int \prod_{j=1}^{v_i} \pi(t_{ij} \mid \lambda_i) \pi(\lambda_i \mid \alpha, \beta) \, d\lambda_i \right]$$
